\chapter{An Introduction to the GSAS-II GUI}
\label{Chap:GSASIIintro}

While this prerequisite chapter is a bit different from the previous, in that I do not expect that you will have previously seen the information here, I do think that having a familiarity with GSAS-II will be helpful before moving on to the rest of the book. In subsequent sections, I will frequently alternate between introducing concepts and then showing how those concepts are embodied when using GSAS-II. For this reason I want to provide an orientation for how GSAS-II is installed and organized here.

\section{Installing GSAS-II \& Web Resources}
GSAS-II is hosted on GitHub.com and is typically downloaded from there. All information related to GSAS-II is indexed via the GitHub home page, \texttt{https://GSASII.github.io}.\index{GSAS-II home page} Please refer to the section of that website (\texttt{https://advancedphotonsource.github.io/GSAS-II-tutorials/install.html}) which provides installation instructions\index{GSAS-II installation instructions}. Most people will choose to download one of the GSAS2MAIN self-installers, but there are many different ways that GSAS-II can be installed depending on your own preferences and goals. The installation instructions web pages provide a detailed  discussion of the options and advantages. It is worth noting that the source code for GSAS-II is installed as part of the GSAS-II installation, so we try to keep the size of that small. This source code is found in GitHub repository  \texttt{https://github.com/AdvancedPhotonSource/GSAS-II}, but there are three other repositories associated with GSAS-II. The GSAS-II documentation, including tutorial web pages, and input used to create  the web pages, help pages and newer tutorials is found at \texttt{https://github.com/AdvancedPhotonSource/GSAS-II-tutorials}. Routines used to compile and package GSAS-II, as well as the distribution files for installation are found in repo https://github.com/AdvancedPhotonSource/GSAS-II-buildtools and finally the files used to create this book are found at \texttt{https://github.com/briantoby/PowderCrystallography}. The GSAS-II code developer's manual, which is a formatted version of text and comments found in the source code is hosted on a separate site, \texttt{https://gsas-ii.readthedocs.io/}.\index{GSAS-II Developer's manual} The documentation section of the home page provides access to the manuals in several formats.  

\section{The GSAS-II GUI}
When initially started, GSAS-II uses three windows. As it is used it will commonly open  "modal" dialogs, which are additional, usually small, windows that "lock" the application until information is entered. It occasionally may open other windows, such as web pages, that work independently of the rest of the program and may persist until that window is closed. The three windows are, the main window, the graphics window and the console window, which will be discussed in the following sections. 
\subsection{The main GSAS-II window}

\subsection{The GSAS-II graphics window}
\subsection{The GSAS-II console window}