\newpage
%\chapter*{Dedication}
\vspace*{\fill}

\begin{center}
{\LARGE\bfseries Dedication}
\end{center}
\vspace{2\baselineskip}
\begin{center}
    
In memory of Sidney Toby,\\
and honoring Diane, Cassie and Josh Toby.
\end{center}
\vspace{\fill}

\newpage
%\chapter*{Thanks}
\begin{center}
{\LARGE\bfseries Thanks}
\end{center}
\vspace{2\baselineskip}

I have benefited from many mentors over my years. I want to thank Joseph Potenza\index{Potenza, J.} and Harvey Schugar\index{Schugar, H.}, my undergraduate advisors at Rutgers, who started my journey in crystallography. Many other members of the Chemistry Department were also great mentors -- too many to mention. Alas, too few who who are still with us. While my subsequent years at Caltech were a time of great personal growth, perhaps the less said of my academic experience there the better, but I did have the opportunity to spend some time with and T.A. for Dick Marsh\index{Marsh, R.}, from whom I learned much; he was always a pleasure to talk with. After Caltech, while employed by Union Carbide, Corp. (rest in pieces), I was introduced to high-resolution powder diffraction, neutron diffraction and Rietveld by David Cox\index{Cox, D.} at Brookhaven National Lab, and first met Takeshi Egami\index{Egami, T.}, then at the University of Pennsylvania. Both changed my life for the better, in ways too great to explain here. I also got to meet Larry Finger\index{Finger, L.} during his experiments at the NSLS. He showed me that one can be a great scientist, as well as a computer hardware and software expert. During my years at Penn working for Takeshi, I learned much about TOF diffraction and its analysis from Jim Jorgenson\index{Jorgenson, J.} and Frank Rotella\index{Rotella, F.} while making measurements Argonne's IPNS neutron source. It was around then that I also started asking questions of Bob Von Dreele\index{Von Dreele, R. B.} and Allen Larson\index{Larson, A.}, then both at LANSCE in Los Alamos. At some point Bob entrusted me with the VAX source code for GSAS and I eventually even had the temerity to provide modified code back. My few years at Air Products were good ones and APCI bought me a \$60K Silicon Graphics workstation that forced me out of my VAX comfort zone. At NIST, I benefited greatly by being able to work with and learn from Judy Stalick \index{Stalick, J.}-- very much missed. The users and staff at the NCNR were a great privilege to work with, but I want to single out Tony Santoro\index{Santoro, A.}, Ted Prince\index{Prince, E.} and Jack Rush\index{Rush, J.}; all no longer with us, alas.

Bob Von Dreele has been a mentor, teacher and work partner, for so many years that I do not know where to begin in thanking him. Working together, we have been recognized for a couple of awards, so it has been quite a partnership. He even had to put up with me as a supervisor twice.
Bob and I had the good fortune to land at the kind shores of Argonne, at a time when we were both in need, and that allowed us to collaborate much more closely. I thank Peter Lee\index{Lee, P.}, Mark Beno\index{Beno, M.} and Gabrielle Long\index{Long, G.} for that. Over the years, I have supervised a few postdocs and interacted with many more, as well as even more students -- they should not think that learning went in only one direction. 

It seems that many members of the crystallographic community marry into the mental health system, and I am no exception. My partner in every aspect of my life, psychologist Diane Pies Toby, Ph.D.,\index{Toby, D. P.} has helped me grow my career in so many ways, not to mention reading and critiquing much of what I write. Our children, Cassie and Josh, have also helped me grow and are always a joy (even when bringing \textit{tsouris}) in my life.

My mother, Frina,\index{Toby, F. S.} introduced me to research computing and my father, Sidney\index{Toby, S.}, showed me how much satisfaction that research and teaching bring. I was actually able to coauthor a paper with both of them during my Rutgers undergrad days. I hope my sister, Carole, has forgiven me for, no doubt, some very boring dinner conversations. 