\chapter{Prerequisite: X-ray and Neutron Scattering}

\section{Terminology: 2$\theta$ and Q}

Diffraction is a quantum phenomenon that requires a tiny particle with a wavelength that is on the order of the size of atoms. Diffraction measurements are routinely done with x-rays (photons) or neutrons and increasingly commonly with electrons, traditionally with electron microscope, but some newer instruments are optimized specifically for diffraction. Scattering of alpha particles or neutral atoms is sometimes performed – but not for crystallography, e.g. structural information on the arrangement of atoms. Also, though not considered here, muon scattering can be valuable for magnetic studies. I will use the term quantum particles to discuss features that x-rays, neutrons, and electrons all share, but it is worth noting that for diffraction studies, energies for neutrons, x-rays and electrons are typically mili-eV, kilo-eV and mega-eV, respectively, to generate wavelengths that are on the order of 1 Å.

 One of the most fundamental concepts in scattering is what physicists call momentum transfer: a measure of the deflection of the diffracted quantum particle from that of the impinging beam. This angle is for historical reasons described as 2$\theta$ and is usually measured in degrees. Thus, forward scattering, where the particle does not have its scattering direction changed, will have a 2$\theta$ value of zero and complete backscattering, where the particle is scattered back towards the source will have 2$\theta$=180°. The 2$\theta$ value itself is not useful without knowing the wavelength of the particle. Further, an important class of diffraction measurements is done with energy-dispersive diffraction, where instead of measuring scattering at a fixed wavelength while varying the angle, the measurements are made at a fixed scattering angle as a function of wavelength. Thus, it is best to get away from angular measurements and use an absolute term that removes the wavelength dependence. Traditionally, crystallographers have used the quantity $\sin\theta/\lambda$ for this, but physicists use a related quantity Q, where Q = $ 2 \pi \sin\theta/\lambda$. I am a chemist by training and self-identify as a crystallographer, but I prefer Q. 
